Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-22 Manuscript/manuscript.tex

The ECC correction was applied to the Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone, and carbonyl regions, which are the most polar parts in the  lipids and
are thus expected to have the largest contribution to the cation binding.
We do not modify the hydrocarbon chain parameters, as they
do not come in direct contact with salt ions and are
already highly optimized to provide a good description for the
hydrophobic part of lipid bilayers \cite{ollila16}.
This is in contrasts with the glycerol backbone and the head group regions, 
which call for improvements in all available lipid models~\cite{botan15}.

@@

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding in lipid bilayers was compared between experiments and simulations by 
using the lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}, which
is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of unit charge bound per lipid~\cite{seelig87}.
The change of order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

@@

The relation between bound charge per lipid,  $X^\pm$, and
the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$,
is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for the binding of
Ca$^{2+}$ to POPC bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because the head group order parameters
decrease with bound positive charge and increase with bound negative
charge~\cite{ollila16,catte16}. This can be rationalized as a change
of the lipid head group dipole tilt toward water phase with the bound positive
charge and {\it vice versa} with the negative charge~\cite{seelig87}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium) in POPC bilayer from simulations
    and experiments \cite{scherer89}.
  }
  \todo{Labeling should be consistent with the previous figure, i.e. experimental data with diamonds.} \\
  \todo{x-axis scale from -1 to 51 to make the point in zero fully visible and to remove empty space.} \\
  \todo{Empty space between figures and from the right column could be reduced.}
\end{figure}

@@

The improved cation binding behaviour to POPC bilayer pave the way for
simulations of complex biochemical systems with correctly described
electrostatic interactions in the vicinity of cellular membranes.
The ECC-lipids model is build by scaling the partial charges and
the LJ-radius of the headgroup, glycerol backbone and carbonyl atoms of
the Lipid14 POPC model~\cite{dickson14}. While the Lipid14 model
is compatible with the AMBER force field family, the compatibility of
the ECC-lipids model may be compromised due to the changes in intermolecular
interactions of the scaled atoms. On the other hand, a fully consistent
ECC-force field should include the correction also in other than lipid
molecules, including water. The work toward this direction and the extension
to other lipid molecules and force fields is left for the future studies.

>>>>

The ECC correction was applied to the Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone, and carbonyl regions, which are the most polar parts in the  lipids and
are thus expected to have the largest contribution to the cation binding.
We do not modify the hydrocarbon chain parameters, as they
do not come in direct contact with salt ions and are
already highly optimized to provide a good description for the
hydrophobic part of lipid bilayers \cite{dickson14,ollila16}.
This is in contrasts with the glycerol backbone and the head group regions, 
which call for improvements in all available lipid models~\cite{botan15}.

@@

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding in lipid bilayers was compared between experiments and simulations by 
using the lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}, which
is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of charge bound per lipid~\cite{seelig87}.
The change of order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

@@

The relation between the amount of the bound charge per lipid,  $X^\pm$, and
the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$,
is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for the binding of
Ca$^{2+}$ to POPC bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because the head group order parameters
decrease with bound positive charge and increase with bound negative
charge~\cite{ollila16,catte16}. This can be rationalized as a change
of the lipid head group dipole tilt toward water phase with the bound positive
charge and {\it vice versa} with the negative charge~\cite{seelig87}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium) in POPC bilayer from simulations
    and experiments \cite{scherer89}.
  }
  \todo{Labeling should be consistent with the previous figure, i.e. experimental data with diamonds.} \\
  \todo{x-axis scale from -1 to 51 to make the point in zero fully visible and to remove empty space.} \\
  \todo{Empty space between figures and from the right column could be reduced.} \\
  \todo{Put the experimental data with larger points and thicker line in this and in the next two figures to
    make more clear separation between simulation and experimental data.}
\end{figure}

@@

The improved cation binding behaviour to POPC bilayer pave the way for
simulations of complex biochemical systems with correctly described
electrostatic interactions in the vicinity of cellular membranes.
The ECC-lipids model is build by scaling the partial charges
and the LJ-radius of the headgroup, glycerol backbone and carbonyl atoms of
the Lipid14 POPC model~\cite{dickson14} with the constant factor
of 0.8 and 0.89, respectively. While the Lipid14 model
is compatible with the AMBER force field family, the compatibility of
the ECC-lipids model may be compromised due to the changes in intermolecular
interactions of the scaled atoms. On the other hand, a fully consistent
ECC-force field should include the correction also in other than lipid
molecules, including water. The work toward this direction and the extension
to other lipid molecules and force fields is left for the future studies.
